---
title: "MLB Hot-Hand Analysis"
author: "Shenwei Zhang"
date: "2026-09-10"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyr)
library(dplyr)
library(lubridate)
library(zoo)
library(ggplot2)
library(readr)
```

# Research Question

# Data and Outcome Construction

## Raw Statcast Data

The analysis uses MLB Statcast pitch-level data for Michael Harris II across the 2022–2025 seasons. Each *row* of the raw dataset represents an *individual pitch* rather than a complete plate appearance or official at-bat.

The dataset contains pitch sequencing variables, plate-appearance outcomes, pitch characteristics, pitch location, and batted-ball information. The main analysis uses the sequencing and outcome variables to reconstruct official at-bats and *create a binary hit indicator*.

Because batting average is defined at the official-at-bat level, the raw pitch-level observations cannot be used directly as the unit of analysis.
The following sections therefore *transform the data* from individual pitches into chronologically ordered plate appearances and official at-bats.

```{r message=FALSE, warning=FALSE}

raw_data <- read_csv("../data/savant_data.csv", show_col_types = FALSE)
# head(raw_data)

tibble(
  observations = nrow(raw_data),
  variables = ncol(raw_data),
  seasons = paste(
    sort(unique(raw_data$game_year)),
    collapse = ", "
  )
)


```

```{r}
raw_data %>%
  select(
    game_date,
    game_year,
    game_pk,
    at_bat_number,
    pitch_number,
    events,
    description
  ) %>%
  slice_head(n = 5)
```


## Chronological Ordering

Because the analysis focuses on sequential batting performance, the ordering of observations must reflect the actual order in which pitches and plate appearances occurred.

The raw Statcast observations were therefore ordered by game date, game identifier, at-bat number, and pitch number. This preserves the within-game and within-plate-appearance sequence required for subsequent
plate-appearance reconstruction and rolling-window analysis.

```{r}
pitch_data <- raw_data %>%
  mutate(game_date = as.Date(game_date,format("%m/%d/%Y")))%>%
  arrange(game_date, game_pk, at_bat_number, pitch_number)

# head(pitch_data)
```
```{r}
class(raw_data$game_date)
head(raw_data$game_date,10)
```


## Plate Appearance Reconstruction

first we create the plate appearance identifier:

```{r}
pitch_data <- pitch_data %>%
  mutate(
    pa_id = paste(game_pk, at_bat_number, sep = "_")
  )

# head(pitch_data)
```

then we need to define a column means the final pitch result to represent this outcome for this plate Appearance..

```{r}
pitch_data <- pitch_data %>%
  group_by(pa_id) %>%
  mutate(final_pitch=pitch_number==max(pitch_number))%>%
  ungroup()
```

now we create a dataset that is based on the PA-level dataset and this only cover the final result for each pa_id, which is count as the current plate Apearance result to define if this is a outs or not.

```{r}
pa_data <- pitch_data %>%
  filter(final_pitch) %>%
  arrange(game_date,game_pk, at_bat_number)
```

## Official At-Bat Definition


Batting average is defined using official at-bats rather than all plate appearances. Therefore, plate appearances that do not count toward the batting-average denominator, such as walks, hit-by-pitches, sacrifice flies,
and catcher interference, were excluded from the official-at-bat sequence.

Plate appearances resulting in hits, strikeouts, field outs, double plays,
fielding errors, or fielder's-choice outcomes were classified as official
at-bats.

This part Our GOAL is: 
not every hit time is a official at-bat. only the 
  “single",
  "double",
  "triple",
  "home_run",
  "field_out",
  "force_out",
  "strikeout",
  "strikeout_double_play",
  "grounded_into_double_play",
  "double_play",
  "triple_play",
  "field_error",
  "fielders_choice",
  "fielders_choice_out"
  count. The walk, hit-by -pitch, sacrifice fly are not count.

```{r}
official_ab_events <- c(
  "single",
  "double",
  "triple",
  "home_run",
  "field_out",
  "force_out",
  "strikeout",
  "strikeout_double_play",
  "grounded_into_double_play",
  "double_play",
  "triple_play",
  "field_error",
  "fielders_choice",
  "fielders_choice_out"
)

pa_data <- pa_data %>%
  mutate(
    official_at_bat = if_else(
      events %in% official_ab_events,
      1L,
      0L
    )
  )
```


SO here we check which kind of event counts as the PA?

```{r}
pa_data %>%
  count(events, official_at_bat, sort = TRUE)
```




## Hit Indicator

define whether it hit or not:



```{r}
pa_data <- pa_data %>%
  mutate(
    hit = if_else(
      events %in% c("single", "double", "triple", "home_run"),
      1L,
      0L
    )
  )


```

A binary hit indicator was defined from the terminal plate-appearance
outcome. Singles, doubles, triples, and home runs were classified as hits
(`hit = 1`); all other plate-appearance outcomes were assigned `hit = 0`.

The hit indicator alone does not determine whether an observation enters
the batting-average denominator. Only plate appearances previously
classified as official at-bats are retained for the primary batting
analysis.


## Official-At-Bat-Level Dataset

The primary analysis dataset was constructed by retaining only plate
appearances classified as official at-bats. Each row of this dataset
therefore represents one chronologically ordered official at-bat with a
binary hit outcome.

This official-at-bat sequence forms the basis for the rolling batting
average and subsequent Monte Carlo streakiness analysis.

```{r}
ab_data <- pa_data %>%
  filter(official_at_bat == 1L) %>%
  arrange(
    game_date,
    game_pk,
    at_bat_number
  )
```


# Descriptive Performance Dynamics

## Season-Level Performance

So here in our *ab_data*, we could garantee the one observation line is a official at-bat.

we want ask some basics questions:
- what is the official AB Michael Harrise II have in each season?
- How many hits?
- what is the season batting average?


```{r}
season_summary <- ab_data %>%
  group_by(game_year) %>%
  summarise(
    official_ab = n(),
    hits = sum(hit),
    batting_average = mean(hit),
    .group = "drop"
  )

season_summary
```

Season-level batting performance was summarized using the official-at-bat
sequence constructed above. For each season, the number of official at-bats,
total hits, and batting average were calculated.

Because the hit indicator is binary, the mean of the hit indicator is
equivalent to the conventional batting average:

$$
\widehat{p}_s
=
\frac{\sum_{i=1}^{n_s} H_{is}}{n_s}
=
\frac{\text{Hits}_s}{\text{Official AB}_s}.
$$

These season-specific batting averages provide a descriptive baseline and
later serve as the estimated hit probabilities under the primary Bernoulli
null model.

## 25-AB Rolling Batting Average

In season s, do we have continuous fluctuation? 

create a variable to record the local performance measure.

here we use the 25 as our window size.

```{r}
window_size <- 25

ab_data <- ab_data %>%
  group_by(game_year) %>%
  arrange(
    game_date,
    game_pk,
    at_bat_number,
    .by_group = TRUE
  ) %>%
  mutate(
    season_ab_number = row_number(),
    rolling_ba_25 = zoo::rollmean(
      hit,
      k = window_size,
      fill = NA,
      align = "right"
    )
  ) %>%
  ungroup()
```


## Observed Performance Patterns


```{r}
ggplot(
  ab_data %>% filter(!is.na(rolling_ba_25)),
  aes(
    x = season_ab_number,
    y = rolling_ba_25
  )
) +
  geom_line(linewidth = 0.8) +
  facet_wrap(
    ~ game_year,
    scales = "free_x"
  ) +
  labs(
    title = "25-At-Bat Rolling Batting Average by Season",
    x = "Official At-Bat Number Within Season",
    y = "25-AB Rolling Batting Average"
  ) +
  theme_minimal()
```

The 25-at-bat rolling batting-average series show substantial local
variation within each season, with periods of relatively high and low
observed performance. The magnitude and persistence of these fluctuations
also appear to differ across seasons.

However, visible variation in a rolling batting-average series is not, by
itself, evidence of a hot-hand effect. Even a sequence of independent hit
outcomes with a constant underlying hit probability can produce apparent
runs of high and low performance by chance.

The inferential question is therefore whether the observed within-season
variation is unusually large relative to what would be expected under an
appropriate no-streakiness null model.



# Statistical Evidence for Streakiness

## Null Hypothesis

The null model represents a season in which Michael Harris II has a
constant underlying probability of recording a hit and official
at-bat outcomes are independent.

For season \(y\), let

\[
\hat p_y =
\frac{\text{Hits}_y}
{\text{Official At-Bats}_y}
\]

denote the observed season batting average.

Under the null model, simulated official at-bat outcomes are generated as

\[
H_{i,y}^{sim}
\overset{iid}{\sim}
\text{Bernoulli}(\hat p_y).
\]

Thus, the baseline hit probability is allowed to differ across seasons,
but remains constant across official at-bats within a season.

The null hypothesis does not imply that the observed sequence should be
smooth. Independent Bernoulli outcomes can still generate consecutive
hits, consecutive non-hits, and temporary periods of high or low rolling
batting average by chance.

The inferential question is therefore whether the observed within-season
variation in rolling batting average is unusually large relative to the
variation generated under this independent Bernoulli benchmark.



## Test Streakiness Statistic

To quantify within-season variation in local batting performance, the
primary streakiness statistic is defined as the variance of the 25-at-bat
rolling batting-average series.

For season \(y\), let \(R_{y,t}\) denote the 25-at-bat rolling batting
average at eligible position \(t\). The observed streakiness statistic is

\[
T_y^{obs}
=
\operatorname{Var}(R_{y,t}).
\]

Larger values of \(T_y^{obs}\) indicate greater dispersion in local
batting performance across the season, corresponding to more pronounced
high and low rolling-performance periods.

The statistic summarizes the magnitude of variation across the full
rolling batting-average series rather than relying only on a single
maximum or minimum. However, a large observed variance does not by itself
establish a hot-hand effect, because substantial rolling variation can
also arise from random sequences of independent hit outcomes.

Statistical evidence is therefore obtained by comparing the observed
variance with the distribution of the same statistic generated under the
season-specific Bernoulli null model.

```{r}
observed_streakiness <- ab_data %>%
  filter(!is.na(rolling_ba_25)) %>%
  group_by(game_year) %>%
  summarise(
    observed_variance = var(rolling_ba_25),
    .groups = "drop"
  )

observed_streakiness
```


## Monte Carlo Simulation

```{r}
simulate_streakiness <- function(
  n_ab,
  p_hit,
  window = 25
) {

  simulated_hits <- rbinom(
    n = n_ab,
    size = 1,
    prob = p_hit
  )

  simulated_rolling_ba <- zoo::rollmean(
    simulated_hits,
    k = window,
    fill = NA,
    align = "right"
  )

  simulated_variance <- var(
    simulated_rolling_ba,
    na.rm = TRUE
  )

  return(simulated_variance)
}
```


```{r}
n_sim <- 5000

set.seed(123)

simulation_results <- season_summary %>%
  select(
    game_year,
    official_ab,
    batting_average
  ) %>%
  tidyr::crossing(
    simulation = seq_len(n_sim)
  ) %>%
  rowwise() %>%
  mutate(
    simulated_variance = simulate_streakiness(
      n_ab = official_ab,
      p_hit = batting_average,
      window = window_size
    )
  ) %>%
  ungroup()
```

The sampling distribution of the streakiness statistic under the null
model is not evaluated analytically. Instead, it is approximated using
Monte Carlo simulation.

For each season, synthetic official-at-bat sequences are generated from

\[
H_{i,y}^{*}
\overset{\mathrm{iid}}{\sim}
\mathrm{Bernoulli}(\hat p_y),
\]

while preserving the observed number of official at-bats and the
season-specific batting average.

For each simulated season, the same 25-at-bat rolling batting-average
procedure used for the observed data is applied. The variance of the
resulting rolling series is then calculated,

\[
T_y^{*}
=
\operatorname{Var}(R_{y,t}^{*}).
\]

This procedure is repeated 5,000 times for each season, producing an
empirical null distribution of the streakiness statistic. These
distributions represent the amount of rolling-performance variation that
would be expected under independent hit outcomes with a constant
within-season hit probability.

## Monte Carlo P-Values

```{r}
monte_carlo_results <- simulation_results %>%
  left_join(
    observed_streakiness,
    by = "game_year"
  ) %>%
  group_by(game_year) %>%
  summarise(
    observed_variance = first(observed_variance),

    exceedances = sum(
      simulated_variance >= observed_variance
    ),

    n_simulations = n(),

    monte_carlo_p = (
      exceedances + 1
    ) / (
      n_simulations + 1
    ),

    .groups = "drop"
  )

monte_carlo_results
```

For each season, the observed streakiness statistic was compared with its
Monte Carlo null distribution. Because larger rolling-variance values
represent greater within-season performance fluctuation, evidence against
the null model is assessed in the upper tail of the simulated distribution.

The Monte Carlo p-value is calculated as

\[
p_{\mathrm{MC}}
=
\frac{
1+\sum_{b=1}^{B}
I(T_{y,b}^{*}\geq T_y^{obs})
}{
B+1
},
\]

where \(B=5{,}000\) is the number of simulated seasons and
\(T_{y,b}^{*}\) is the simulated streakiness statistic from Monte Carlo
replication \(b\).

The finite-simulation correction prevents a p-value of exactly zero when
none of the simulated statistics is as large as the observed statistic.

Small Monte Carlo p-values indicate that the observed rolling-performance
variation is unusual relative to the season-specific i.i.d. Bernoulli
null model. They should not be interpreted as the probability that the
null hypothesis is true or as direct proof of a causal hot-hand mechanism.



## Null-Envelope Visualization

to see which part of the data is the extreme place?

```{r}
null_cutoffs <- simulation_results %>%
  group_by(game_year) %>%
  summarise(
    null_95 = quantile(
      simulated_variance,
      probs = 0.95
    ),
    .groups = "drop"
  )

null_cutoffs
```

```{r}
null_plot_data <- null_cutoffs %>%
  left_join(
    observed_streakiness,
    by = "game_year"
  )

null_plot_data
```

```{r}
ggplot(
  simulation_results,
  aes(x = simulated_variance)
) +
  geom_histogram(
    bins = 40
  ) +
  geom_vline(
    data = null_plot_data,
    aes(xintercept = null_95),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_vline(
    data = null_plot_data,
    aes(xintercept = observed_variance),
    inherit.aes = FALSE,
    linewidth = 1
  ) +
  facet_wrap(
    ~ game_year,
    scales = "free"
  ) +
  labs(
    title = "Observed Streakiness Relative to the Monte Carlo Null Distribution",
    subtitle = "Solid line = observed variance; dashed line = 95th percentile of the null distribution",
    x = "Variance of 25-AB Rolling Batting Average",
    y = "Number of Simulated Seasons"
  ) +
  theme_minimal()
```

```{r}
ggplot(
  simulation_results,
  aes(x = simulated_variance)
) +
  geom_rect(
    data = null_plot_data,
    aes(
      xmin = null_95,
      xmax = Inf,
      ymin = -Inf,
      ymax = Inf
    ),
    inherit.aes = FALSE,
    fill = "#FEE0D2",
    alpha = 0.45
  ) +
  geom_histogram(
    bins = 40,
    fill = "#6BAED6",
    color = "white",
    alpha = 0.8
  ) +
  geom_vline(
    data = null_plot_data,
    aes(xintercept = null_95),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 0.8,
    color = "#636363"
  ) +
  geom_vline(
    data = null_plot_data,
    aes(xintercept = observed_variance),
    inherit.aes = FALSE,
    linewidth = 1,
    color = "#CB181D"
  ) +
  facet_wrap(
    ~ game_year,
    scales = "free"
  ) +
  labs(
    title = "Observed Streakiness Relative to the Monte Carlo Null Distribution",
    subtitle = "Red line = observed variance; dashed line = 95th percentile; shaded region = upper 5% of null distribution",
    x = "Variance of 25-AB Rolling Batting Average",
    y = "Number of Simulated Seasons"
  ) +
  theme_minimal()
```

The Monte Carlo null distributions provide a visual benchmark for the
amount of rolling-performance variation expected under independent
Bernoulli outcomes.

The 2022 observed statistic lies well within its null distribution, while
the 2023 statistic is farther into the upper tail but remains below the
95th-percentile cutoff. The 2024 observed variance lies close to the
upper-tail threshold, consistent with suggestive but not conventionally
significant evidence of elevated variation.

The strongest departure occurs in 2025. The observed rolling-BA variance
lies beyond the 95th-percentile cutoff and exceeds all 5,000 simulated
null statistics. This is consistent with the corrected Monte Carlo
p-value of approximately 0.0002.

These comparisons indicate that the magnitude of rolling-performance
variation in 2025 is highly unusual relative to the specified
season-specific i.i.d. Bernoulli null model. They do not, by themselves,
identify the mechanism responsible for that departure.

## Pointwise Null Envelope


```{r}
# simulated rolling path

simulate_rolling_path <- function(
  n_ab,
  p_hit,
  window = 25
) {

  simulated_hits <- rbinom(
    n = n_ab,
    size = 1,
    prob = p_hit
  )

  simulated_rolling_ba <- zoo::rollmean(
    simulated_hits,
    k = window,
    fill = NA,
    align = "right"
  )

  tibble(
    season_ab_number = seq_len(n_ab),
    simulated_rolling_ba = simulated_rolling_ba
  )
}
```

```{r}
# generate simulations

set.seed(123)

rolling_simulations <- season_summary %>%
  select(
    game_year,
    official_ab,
    batting_average
  ) %>%
  tidyr::crossing(
    simulation = seq_len(n_sim)
  ) %>%
  rowwise() %>%
  mutate(
    rolling_path = list(
      simulate_rolling_path(
        n_ab = official_ab,
        p_hit = batting_average,
        window = window_size
      )
    )
  ) %>%
  ungroup() %>%
  select(
    game_year,
    simulation,
    rolling_path
  ) %>%
  tidyr::unnest(rolling_path)
```


```{r}
null_envelope <- rolling_simulations %>%
  filter(!is.na(simulated_rolling_ba)) %>%
  group_by(
    game_year,
    season_ab_number
  ) %>%
  summarise(
    lower_95 = quantile(
      simulated_rolling_ba,
      probs = 0.025
    ),
    upper_95 = quantile(
      simulated_rolling_ba,
      probs = 0.975
    ),
    .groups = "drop"
  )
```

```{r}
envelope_plot_data <- ab_data %>%
  filter(!is.na(rolling_ba_25)) %>%
  select(
    game_year,
    season_ab_number,
    rolling_ba_25
  ) %>%
  left_join(
    null_envelope,
    by = c(
      "game_year",
      "season_ab_number"
    )
  )
```

```{r}
ggplot(
  envelope_plot_data,
  aes(
    x = season_ab_number,
    y = rolling_ba_25
  )
) +
  geom_ribbon(
    aes(
      ymin = lower_95,
      ymax = upper_95
    ),
    alpha = 0.25
  ) +
  geom_line(
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ game_year,
    scales = "free_x"
  ) +
  labs(
    title = "Observed 25-AB Rolling Batting Average vs. Null Envelope",
    subtitle = "Shaded region = pointwise 95% Monte Carlo null envelope",
    x = "Official At-Bat Number Within Season",
    y = "25-AB Rolling Batting Average"
  ) +
  theme_minimal()
```

To visualize where the observed rolling batting-average trajectory differs
from patterns commonly generated under the null model, pointwise Monte
Carlo envelopes were constructed.

For each season and each eligible official-at-bat position, the 2.5th and
97.5th percentiles of the simulated 25-at-bat rolling batting averages
were calculated. These quantiles form a pointwise 95% null envelope around
the rolling-performance trajectories expected under the season-specific
i.i.d. Bernoulli model.

Observed values outside the envelope identify local periods that appear
unusual relative to the null benchmark. However, the envelope is
pointwise rather than simultaneous. Because many positions are examined
within each season, an individual crossing should not be interpreted as
a standalone 5% significance test.

Formal season-level inference therefore remains based on the Monte Carlo
variance statistic and its corresponding p-value. The pointwise envelope
is used primarily to visualize and localize the performance patterns
underlying the season-level results.

## Year-by-Year Results

The Monte Carlo analysis reveals substantial differences in the magnitude
of within-season rolling-performance variation across seasons.

In 2022, the observed rolling-BA variance was well within the range
expected under the season-specific Bernoulli null model
(\(p_{\mathrm{MC}} = 0.634\)). Thus, the observed variation provides
little evidence of a departure from the independent, constant-probability
benchmark.

The 2023 season exhibited greater rolling-performance variation, but the
observed statistic remained compatible with the null distribution
(\(p_{\mathrm{MC}} = 0.138\)).

The 2024 observed variance was farther into the upper tail of its null
distribution (\(p_{\mathrm{MC}} = 0.089\)). This provides suggestive
evidence of elevated within-season variation, although the result does not
cross the conventional 0.05 threshold.

The strongest departure occurred in 2025. None of the 5,000 simulated
seasons produced a rolling-BA variance as large as the observed value.
After applying the finite-simulation correction, the Monte Carlo p-value
was approximately \(0.0002\). The magnitude of the observed 2025
rolling-performance variation was therefore highly unusual relative to
the specified season-specific i.i.d. Bernoulli null model.

These results provide evidence about departures from the specified
no-streakiness benchmark rather than establishing a causal hot-hand
mechanism. Time-varying opponent quality, pitcher matchups, injuries,
platoon effects, and other contextual factors could also generate
departures from a constant-probability Bernoulli model.


Because four season-specific tests were conducted, Holm-adjusted p-values
were also examined as a multiple-testing sensitivity check.
```{r}
monte_carlo_results <- monte_carlo_results %>%
  mutate(
    holm_p = p.adjust(
      monte_carlo_p,
      method = "holm"
    )
  )

monte_carlo_results
```



# Main Findings

The analysis yields three main findings.

First, Michael Harris II's 25-at-bat rolling batting average exhibits
substantial local variation across all four seasons. Periods of relatively
high and low observed performance are therefore visible in the raw
performance trajectories, although such patterns can also arise from
random sequences of independent hit outcomes.

Second, the magnitude of this variation differs substantially across
seasons when compared with season-specific Monte Carlo null distributions.
The 2022 and 2023 observed statistics were compatible with the independent
Bernoulli benchmark, while the 2024 statistic was farther into the upper
tail of its null distribution (\(p_{\mathrm{MC}} = 0.089\)).

Third, the strongest departure occurred in 2025. None of the 5,000
simulated null seasons produced a 25-at-bat rolling-BA variance as large
as the observed value, yielding a finite-simulation-corrected Monte Carlo
p-value of approximately \(0.0002\).

Taken together, the results indicate that the 2025 batting sequence
contained substantially more rolling-performance variation than expected
under the specified season-specific i.i.d. Bernoulli model. This finding
identifies an unusually variable performance pattern, but does not by
itself establish a causal hot-hand mechanism.


# Limitations

The analysis should be interpreted relative to the assumptions of the
specified null model.

First, the season-specific i.i.d. Bernoulli model assumes that the
underlying hit probability remains constant within each season and that
official at-bat outcomes are independent. Real batting opportunities vary
with pitcher quality, handedness, game environment, injuries, and other
contextual factors. Consequently, rejection of this simplified null model
does not uniquely identify a hot-hand mechanism.

Second, the season-level hit probability is unknown and is replaced by the
observed batting average. The Monte Carlo distributions therefore
condition on an estimated season-specific probability rather than
propagating uncertainty in that parameter.

Third, the primary analysis depends on a 25-at-bat rolling window and on
rolling-BA variance as the streakiness statistic. The window length is an
analytical choice, and variance measures the magnitude of local performance
variation rather than temporal persistence directly. Alternative window
lengths and streakiness statistics should therefore be examined as
robustness checks.

Fourth, adjacent rolling batting averages share most of their underlying
at-bats and are consequently highly dependent. The Monte Carlo procedure
reproduces this overlapping-window structure in every simulated season,
but individual rolling observations should not be interpreted as
independent measurements.

Finally, the 95% rolling-BA null envelopes are pointwise rather than
simultaneous. Individual crossings are therefore descriptive indicators
of locally unusual performance rather than standalone significance tests.

More broadly, the analysis is observational. It identifies departures
from a specified no-streakiness statistical benchmark but does not
estimate a randomized or causal effect of being in a "hot" state.

# Conclusion

This study examined whether Michael Harris II's within-season batting
performance exhibited more temporal variation than would be expected under
a season-specific no-streakiness benchmark. Pitch-level Statcast data from
2022 through 2025 were reconstructed into official at-bats, represented as
binary hit outcomes, and summarized using 25-at-bat rolling batting
averages. The variance of each season's rolling series was then compared
with empirical null distributions generated from 5,000 season-specific
Bernoulli simulations.

The strength of evidence differed substantially across seasons. The 2022
and 2023 performance trajectories were broadly compatible with the
independent Bernoulli benchmark, while 2024 showed suggestive evidence of
elevated rolling-performance variation. The strongest departure occurred
in 2025: none of the 5,000 simulated null seasons produced a rolling-BA
variance as large as the observed value, corresponding to a
finite-simulation-corrected Monte Carlo p-value of approximately 0.0002.

These findings indicate that Harris's 2025 batting sequence contained
substantially more local performance variation than expected under the
specified season-specific i.i.d. Bernoulli model. They do not, however,
establish that the variation was generated by a causal or persistent
hot-hand mechanism. Robustness analyses and baseball-context analyses are
therefore needed to determine whether the result persists under alternative
statistical specifications and whether changes in pitching environment,
swing decisions, contact quality, or other observable factors coincide
with the unusual performance periods.
