---
title: "Robustness Analysis: MLB Hot-Hand Effect"
author: "Shenwei Zhang"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---


```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(zoo)
```


# Purpose

The primary analysis identified unusually large within-season variation
in Michael Harris II's 2025 batting performance relative to a
season-specific i.i.d. Bernoulli null model. That analysis used a
25-at-bat rolling batting average and the variance of the rolling series
as the primary streakiness statistic.

These choices define one reasonable statistical specification, but they
are not unique. The purpose of this robustness analysis is therefore to
evaluate whether the substantive findings persist under alternative
reasonable analytical choices.

The analysis focuses on three sources of specification sensitivity:

1. the length of the rolling batting-average window;
2. the statistic used to quantify temporal variation or dependence; and
3. the assumptions used to construct the no-streakiness benchmark.

The goal is not to search across specifications for statistically
significant results. Instead, the alternative specifications are used as
sensitivity checks for conclusions established in the primary analysis.

# Reconstruct Analysis Data

```{r}
raw_data <- readr::read_csv(
  "../data/savant_data.csv",
  show_col_types = FALSE
)

```


## Load Statcast Data

```{r}
source("../data/prepare_analysis_data.R")

dim(ab_data)
```


## Construct Official At-Bat Dataset

```{r}
ab_data %>%
  group_by(game_year) %>%
  summarise(
    official_ab = n(),
    hits = sum(hit),
    batting_average = mean(hit),
    .groups = "drop"
  )
```

# Rolling-Window Sensitivity



## Alternative Window Lengths

The primary analysis uses a 25-at-bat rolling window to summarize local
batting performance. Because the choice of window length determines the
time scale over which performance variation is measured, the analysis is
repeated using several alternative window lengths.

The sensitivity analysis considers windows of 10, 15, 25, and 30 official
at-bats:

\[
w \in \{10, 15, 25, 30\}.
\]

```{r}
window_sizes <- c(10, 15, 25, 30)

window_sizes
```


The 25-at-bat window remains the primary specification. The alternative
windows are used only to evaluate whether the substantive conclusions are
sensitive to the chosen smoothing scale.



## Observed Rolling Patterns


For each window length, rolling batting average is recalculated separately
within each season. Streakiness is measured using the same statistic as
the primary analysis: the variance of the rolling batting-average series,

\[
T_y(w) = \operatorname{Var}\left(R_{y,t}^{(w)}\right).
\]

Keeping the statistic fixed while changing only the window length allows
the sensitivity of the observed variation to the smoothing choice to be
evaluated directly.


```{r}
calculate_observed_streakiness <- function(
  data,
  window
) {

  data %>%
    group_by(game_year) %>%
    arrange(
      game_date,
      game_pk,
      at_bat_number,
      .by_group = TRUE
    ) %>%
    mutate(
      rolling_ba = zoo::rollmean(
        hit,
        k = window,
        fill = NA,
        align = "right"
      )
    ) %>%
    summarise(
      observed_variance = var(
        rolling_ba,
        na.rm = TRUE
      ),
      .groups = "drop"
    ) %>%
    mutate(
      window = window
    )
}
```






```{r}
observed_window_results <- bind_rows(
  lapply(
    window_sizes,
    function(w) {
      calculate_observed_streakiness(
        data = ab_data,
        window = w
      )
    }
  )
)

observed_window_results
```

Observed rolling-BA variance changes with the window length, as expected
because shorter windows are more responsive to local fluctuations while
longer windows smooth the performance series.

Across the specifications, 2025 consistently shows relatively large
observed variation. However, raw variances are not directly comparable
across window lengths because each window produces a different scale of
natural variability. Statistical interpretation therefore requires a
separate Monte Carlo null distribution for each season-window combination.

## Monte Carlo Tests Across Windows

**For each window, is the observed variance unusually large relative to what that same window would produce under the Bernoulli null?**

### create the simulation function

Raw rolling-BA variances are not directly comparable across window lengths.
Therefore, a separate Monte Carlo null distribution is constructed for
each season-window combination.

For each simulation, a season-length sequence of independent Bernoulli
outcomes is generated using the observed season batting average. The same
rolling window and variance statistic are then applied to the simulated
sequence, ensuring that each observed statistic is compared with an
appropriate null benchmark.

```{r}
simulate_streakiness <- function(
  n_ab,
  p_hit,
  window
) {

  simulated_hits <- rbinom(
    n = n_ab,
    size = 1,
    prob = p_hit
  )

  simulated_rolling_ba <- zoo::rollmean(
    simulated_hits,
    k = window,
    fill = NA,
    align = "right"
  )

  var(
    simulated_rolling_ba,
    na.rm = TRUE
  )
}
```

### prepare the simulation parameters for each season

Each simulated season preserves the observed number of official at-bats \(n_y\) and season batting average \(\hat p_y\). Hit outcomes are then independently generated from \(\text{Bernoulli}(\hat p_y)\), preserving the season's baseline performance while removing the observed temporal structure.

```{r}
season_summary <- ab_data %>%
  group_by(game_year) %>%
  summarise(
    official_ab = n(),
    hits = sum(hit),
    batting_average = mean(hit),
    .groups = "drop"
  )

# season_summary (same as the mainanalysis infor about this)
# since we only change the window size not the other
```

### Generate Monte Carlo Simulations

now this part code create all th combinations: 
$$4 seasons×4 windows×5000 simulations=80,000$$

For each season-window combination, 5,000 Monte Carlo seasons are
generated under the season-specific Bernoulli null model.

The observed and simulated sequences are analyzed using the same rolling window and variance statistic. The observed statistic \(T_y^{obs}(w)\) is then compared with the simulated null statistics \(T_y^*(w)\) to determine whether the observed variation is unusually large under the no-streakiness benchmark.

```{r}
n_sim <- 5000

set.seed(123)

window_simulations <- tidyr::crossing(
  season_summary,
  window = window_sizes,
  simulation = seq_len(n_sim)
) %>%
  rowwise() %>%
  mutate(
    simulated_variance = simulate_streakiness(
      n_ab = official_ab,
      p_hit = batting_average,
      window = window
    )
  ) %>%
  ungroup()
```

```{r}
dim(window_simulations)
# check the simaltion numbers?
```

## Window-Sensitivity Results

For each season-window combination, the observed rolling-BA variance is
compared with its corresponding Monte Carlo null distribution. The
Monte Carlo p-value is the proportion of simulated statistics at least as
large as the observed statistic, using the finite-simulation correction:

\[
p_{\mathrm{MC}}
=
\frac{1+\sum_{b=1}^{B} I(T_{y,b}^{*}(w)\geq T_y^{obs}(w))}
{B+1}.
\]

### calculate the 16-values

```{r}
window_robustness_results <- window_simulations %>%
  left_join(
    observed_window_results,
    by = c("game_year", "window")
  ) %>%
  group_by(
    game_year,
    window
  ) %>%
  summarise(
    observed_variance = first(observed_variance),
    exceedances = sum(
      simulated_variance >= observed_variance
    ),
    n_simulations = n(),
    monte_carlo_p = (
      exceedances + 1
    ) / (
      n_simulations + 1
    ),
    .groups = "drop"
  )

window_robustness_results
```

The 2025 result is robust to the choice of rolling-window length. Across
10-, 15-, 25-, and 30-at-bat windows, the observed rolling-BA variance
remains in the extreme upper tail of its corresponding Monte Carlo null
distribution, with Monte Carlo p-values ranging from approximately
0.0002 to 0.0010.

In contrast, the 2022–2024 results are substantially less extreme across
the tested window lengths. Thus, the unusually large 2025 variation is
not driven specifically by the primary 25-at-bat window choice.

### visualize window sensitivity

```{r}
ggplot(
  window_robustness_results,
  aes(
    x = window,
    y = monte_carlo_p,
    group = factor(game_year),
    linetype = factor(game_year),
    shape = factor(game_year)
  )
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  geom_hline(
    yintercept = 0.05,
    linetype = "dashed"
  ) +
  scale_x_continuous(
    breaks = window_sizes
  ) +
  labs(
    title = "Monte Carlo P-Values Across Rolling-Window Lengths",
    subtitle = "Dashed horizontal line indicates p = 0.05",
    x = "Rolling Window Length (Official At-Bats)",
    y = "Monte Carlo P-Value",
    linetype = "Season",
    shape = "Season"
  ) +
  theme_minimal()
```

The window-sensitivity analysis shows a clear difference across seasons.
The 2022 results remain well within the range expected under the null
model, while the 2023 and 2024 results move closer to the upper tail as
the window length changes but remain above the 0.05 reference level.

In contrast, the 2025 Monte Carlo p-values remain approximately 0.001 or
smaller across all four window lengths. They appear nearly flat at zero
in the figure because the y-axis extends to approximately 0.8, not because
the p-values are actually zero.

Overall, the unusually large rolling-performance variation observed in
2025 is robust to the choice of 10-, 15-, 25-, or 30-at-bat windows and
is therefore not specific to the primary 25-AB specification.


# Alternative Streakiness Statistics

**Does the 2025 result persist when streakiness is measured in a different way?**

## Why Variance May Be Insufficient

The primary analysis measures streakiness using the variance of the
rolling batting-average series. This captures the overall magnitude of
local performance fluctuations, but it does not directly measure how
batting outcomes are ordered over time.

Two seasons can have similar rolling-BA variance while exhibiting
different temporal patterns. Therefore, alternative statistics are used
to test whether the 2025 finding also appears when streakiness is defined
through temporal dependence rather than only the magnitude of rolling
variation.


## Alternative Statistics 1： Lag-1 Autocorrelation

As an alternative to rolling-BA variance, lag-1 autocorrelation measures
whether consecutive official at-bat outcomes are temporally related.

For the binary hit sequence \(H_1,\ldots,H_n\), the statistic is

\[
T_{\text{ACF}}
=
\operatorname{Cor}(H_t,H_{t-1}).
\]

Positive values indicate that similar outcomes tend to occur consecutively,
while values near zero indicate little linear dependence between adjacent
at-bats. This provides a complementary measure of streakiness that focuses
on temporal ordering rather than the magnitude of rolling-BA variation.

### observed lag-1 auto correlation

```{r}
calculate_lag1 <- function(data) {

  data %>%
    group_by(game_year) %>%
    arrange(
      game_date,
      game_pk,
      at_bat_number,
      .by_group = TRUE
    ) %>%
    summarise(
      observed_lag1 = cor(
        hit[-1],
        hit[-length(hit)]
      ),
      .groups = "drop"
    )
}
```

```{r}
observed_lag1_results <- calculate_lag1(ab_data)

observed_lag1_results
```
### monte carlo null distribution for lag-1

**Does the observed lag-1 correlation in 2025 count as the unusually large?**

To isolate sensitivity to the choice of streakiness statistic, the
season-specific i.i.d. Bernoulli null model is kept unchanged from the
primary analysis.

For each simulated season,

\[
H_{i,y}^{*}
\overset{iid}{\sim}
\operatorname{Bernoulli}(\hat p_y),
\]

where the observed number of official at-bats and season batting average
are preserved. Lag-1 autocorrelation is then calculated for each simulated
hit sequence and compared with the observed statistic.

```{r}
simulate_lag1 <- function(
  n_ab,
  p_hit
) {

  simulated_hits <- rbinom(
    n = n_ab,
    size = 1,
    prob = p_hit
  )

  cor(
    simulated_hits[-1],
    simulated_hits[-length(simulated_hits)]
  )
}
```

### Monte Carlo Simulation for Lag-1 Autocorrelation

For each season, 5,000 independent hit sequences are generated under the
season-specific Bernoulli null model. Each simulated season preserves the
observed number of official at-bats and batting average.

$$
4 seasons * 5000 simulations = 20000
$$

Lag-1 autocorrelation is calculated for every simulated sequence, producing
a null distribution that represents the amount of adjacent dependence that
could arise by chance under independent batting outcomes.

```{r}
set.seed(123)

lag1_simulations <- tidyr::crossing(
  season_summary,
  simulation = seq_len(n_sim)
) %>%
  rowwise() %>%
  mutate(
    simulated_lag1 = simulate_lag1(
      n_ab = official_ab,
      p_hit = batting_average
    )
  ) %>%
  ungroup()
```

```{r}
# check it
dim(lag1_simulations)
lag1_simulations %>%
  count(game_year)
```

### Monte Carlo Results

For each season, the observed lag-1 autocorrelation is compared with the
distribution of lag-1 autocorrelations generated under the i.i.d.
Bernoulli null model.

Because the hot-hand hypothesis corresponds to positive temporal
dependence, the upper-tail Monte Carlo p-value is

\[
p_{\mathrm{MC}}
=
\frac{
1+\sum_{b=1}^{B}
I(T_{\mathrm{ACF},b}^{*}\geq T_{\mathrm{ACF}}^{obs})
}{
B+1
}.
\]

A small p-value indicates that the observed positive lag-1 dependence
would be unusual under independent batting outcomes.

```{r}
lag1_results <- lag1_simulations %>%
  left_join(
    observed_lag1_results,
    by = "game_year"
  ) %>%
  group_by(game_year) %>%
  summarise(
    observed_lag1 = first(observed_lag1),

    exceedances = sum(
      simulated_lag1 >= observed_lag1
    ),

    n_simulations = n(),

    monte_carlo_p = (
      exceedances + 1
    ) / (
      n_simulations + 1
    ),

    .groups = "drop"
  )

lag1_results
```
Lag-1 autocorrelation produces a different pattern from the rolling-variance
statistic. The 2025 season has a positive observed lag-1 correlation
(\(r = 0.045\)), but the corresponding Monte Carlo p-value is approximately
0.124. Thus, the unusually large rolling-performance variation identified
for 2025 does not appear to be driven by unusually strong immediate
at-bat-to-at-bat dependence.

In contrast, 2023 exhibits the largest observed lag-1 autocorrelation
(\(r = 0.097\)) and a relatively small upper-tail Monte Carlo p-value
(\(p \approx 0.012\)). This illustrates that different streakiness
statistics capture different aspects of temporal structure: rolling
variance measures the magnitude of local performance fluctuations,
whereas lag-1 autocorrelation measures dependence between consecutive
at-bat outcomes.

## Alternative Statistic 2 — Longest Hit Run

As a second alternative measure, the longest hit run captures the maximum
number of consecutive official at-bats resulting in hits within a season.

\[
T_{\text{run}}
=
\max\{\text{length of consecutive hit sequences}\}.
\]

Unlike rolling-BA variance, which measures the magnitude of local
performance fluctuations, and lag-1 autocorrelation, which measures
average dependence between adjacent outcomes, the longest-run statistic
focuses on extreme clustering of consecutive hits.

```{r}
# define the statistic
calculate_longest_hit_run <- function(hit_sequence) {

  runs <- rle(hit_sequence)

  max(
    runs$lengths[runs$values == 1]
  )
}
```

```{r}
# use it in the real 2022-2025 data
observed_run_results <- ab_data %>%
  group_by(game_year) %>%
  arrange(
    game_date,
    game_pk,
    at_bat_number,
    .by_group = TRUE
  ) %>%
  summarise(
    observed_longest_run =
      calculate_longest_hit_run(hit),
    .groups = "drop"
  )

observed_run_results
```

### Monte Carlo simulate longest hit runs

The observed longest hit run increases from four consecutive hits in 2022
to six in 2025. However, the raw run length alone does not indicate
whether such clustering is unusual, because seasons differ in both the
number of official at-bats and baseline batting average.

Therefore, the observed longest run is compared with season-specific
Monte Carlo null distributions generated under the same i.i.d. Bernoulli
model used in the primary analysis.

```{r}
simulate_longest_hit_run <- function(
  n_ab,
  p_hit
) {

  simulated_hits <- rbinom(
    n = n_ab,
    size = 1,
    prob = p_hit
  )

  calculate_longest_hit_run(
    simulated_hits
  )
}
```


$$
4 sesons * 5000 = 20000
$$

```{r}
set.seed(123)

run_simulations <- tidyr::crossing(
  season_summary,
  simulation = seq_len(n_sim)
) %>%
  rowwise() %>%
  mutate(
    simulated_longest_run =
      simulate_longest_hit_run(
        n_ab = official_ab,
        p_hit = batting_average
      )
  ) %>%
  ungroup()
```

```{r}
run_simulations %>%
  count(game_year)
```

### Monte Carlo Results for Longest Hit Run

For each season, the observed longest hit run is compared with the
distribution of longest runs generated under the i.i.d. Bernoulli null
model.

The upper-tail Monte Carlo p-value is

\[
p_{\mathrm{MC}}
=
\frac{
1+\sum_{b=1}^{B}
I(T_{\mathrm{run},b}^{*}\geq T_{\mathrm{run}}^{obs})
}{
B+1
}.
\]

A small p-value indicates that a hit run at least as long as the observed
run would be unusual under independent batting outcomes.

```{r}
run_results <- run_simulations %>%
  left_join(
    observed_run_results,
    by = "game_year"
  ) %>%
  group_by(game_year) %>%
  summarise(
    observed_longest_run =
      first(observed_longest_run),

    exceedances = sum(
      simulated_longest_run >=
        observed_longest_run
    ),

    n_simulations = n(),

    monte_carlo_p = (
      exceedances + 1
    ) / (
      n_simulations + 1
    ),

    .groups = "drop"
  )

run_results
```
### Summary

The longest-hit-run statistic provides weaker evidence of unusual
consecutive-hit clustering than the primary rolling-variance analysis.
The longest observed run increased from four hits in 2022 to six hits in
2025, but a run of at least six hits occurred in 498 of the 5,000
simulated 2025 seasons, corresponding to a Monte Carlo p-value of
approximately 0.100.

Together with the lag-1 autocorrelation results, this suggests that the
strong 2025 rolling-variance result is not explained simply by unusually
strong immediate hit-to-hit dependence or by a single exceptionally long
run of consecutive hits. Instead, the evidence appears more consistent
with unusual variation in performance over a broader time scale.

# Null-Model Sensitivity

## Limitations of Constant-p Bernoulli Null

The primary analysis uses a season-specific i.i.d. Bernoulli null model,

\[
H_{i,y}^{*}
\overset{iid}{\sim}
\operatorname{Bernoulli}(\hat p_y),
\]

which assumes that every official at-bat within a season has the same hit
probability and that outcomes are independent over time.

This provides a transparent benchmark for no temporal structure, but it
simplifies the actual baseball data-generating process. In observational
game data, the probability of a hit may vary across at-bats because of
changing game contexts, opponents, pitch characteristics, and other
observed or unobserved factors.

Therefore, an unusually large statistic relative to the i.i.d. Bernoulli
null establishes that the observed temporal pattern is inconsistent with
this particular constant-probability benchmark. It does not, by itself,
identify a causal hot-hand effect.

The following sensitivity analyses therefore consider alternative null
models that preserve more structure from the observed data while removing
the specific temporal ordering associated with streakiness.

## Alternative Benchmarks

### Within-Season Permutation Null

As an alternative to the i.i.d. Bernoulli benchmark, a within-season
permutation null preserves the exact observed hit and non-hit outcomes
within each season while randomly reordering their temporal positions.

For season \(y\), each permutation therefore preserves the number of
official at-bats and the total number of hits exactly, but removes the
observed ordering of those outcomes.

This null asks whether the observed rolling-performance variation is
unusual conditional on the season's realized outcomes, rather than
relative to newly generated Bernoulli outcomes. The permutation procedure
is used as a statistical null benchmark and should not be interpreted as
a randomized experimental assignment mechanism.

```{r}
simulate_permutation_variance <- function(
  hit_sequence,
  window = 25
) {

  permuted_hits <- sample(
    hit_sequence,
    size = length(hit_sequence),
    replace = FALSE
  )

  permuted_rolling_ba <- zoo::rollmean(
    permuted_hits,
    k = window,
    fill = NA,
    align = "right"
  )

  var(
    permuted_rolling_ba,
    na.rm = TRUE
  )
}

```

#### prepare the observed hit sequences

```{r}
season_sequences <- ab_data %>%
  arrange(
    game_year,
    game_date,
    game_pk,
    at_bat_number
  ) %>%
  group_by(game_year) %>%
  summarise(
    hit_sequence = list(hit),
    .groups = "drop"
  )

season_sequences
```

#### Generate 5000 permutations per season

For each season, the observed hit sequence is randomly permuted 5,000
times. Each permutation preserves the exact number of hits and non-hits
but changes their temporal ordering.

For every permuted sequence, the primary 25-at-bat rolling batting
average and its variance are recalculated. This produces a
season-specific permutation null distribution for the primary
streakiness statistic.

```{r}
set.seed(123)

permutation_simulations <- tidyr::crossing(
  season_sequences,
  simulation = seq_len(n_sim)
) %>%
  rowwise() %>%
  mutate(
    permuted_variance =
      simulate_permutation_variance(
        hit_sequence = hit_sequence,
        window = 25
      )
  ) %>%
  ungroup()
```

```{r}
# check it
permutation_simulations %>%
  count(game_year)

test_sequence <- season_sequences %>%
  filter(game_year == 2025) %>%
  pull(hit_sequence) %>%
  .[[1]]

test_permutation <- sample(
  test_sequence,
  replace = FALSE
)

sum(test_sequence)
sum(test_permutation)

```

#### permutation simulate test results

For each season, the observed 25-at-bat rolling-BA variance is compared
with the distribution obtained from 5,000 random permutations of the
observed hit sequence.

The permutation p-value is

\[
p_{\mathrm{perm}}
=
\frac{
1+\sum_{b=1}^{B}
I(T_{y,b}^{\mathrm{perm}}\geq T_y^{obs})
}{
B+1
}.
\]

A small permutation p-value indicates that the observed temporal
arrangement of hits produces greater rolling-performance variation than
would typically occur under random reorderings of the same season's
observed outcomes.

```{r}
observed_primary_variance <- observed_window_results %>%
  filter(window == 25) %>%
  select(
    game_year,
    observed_variance
  )

```

```{r}
permutation_results <- permutation_simulations %>%
  left_join(
    observed_primary_variance,
    by = "game_year"
  ) %>%
  group_by(game_year) %>%
  summarise(
    observed_variance =
      first(observed_variance),

    exceedances = sum(
      permuted_variance >=
        observed_variance
    ),

    n_permutations = n(),

    permutation_p = (
      exceedances + 1
    ) / (
      n_permutations + 1
    ),

    .groups = "drop"
  )

permutation_results
```

The permutation benchmark produces conclusions similar to the primary
i.i.d. Bernoulli analysis. In 2025, only 6 of 5,000 random reorderings
produced a 25-at-bat rolling-BA variance at least as large as the observed
value of 0.01455, corresponding to a corrected permutation p-value of
approximately 0.0014.

The 2022--2024 seasons were substantially less extreme under the
permutation benchmark, with permutation p-values of approximately 0.648,
0.147, and 0.094, respectively.

Thus, the unusually large 2025 rolling-performance variation is not
explained solely by the Bernoulli null allowing the total number of hits
to vary across simulations. The result persists when the realized number
of hits is held fixed and only their temporal ordering is randomized.

However, the permutation benchmark still treats the temporal positions
of the observed outcomes as exchangeable under the null and does not
account for time-varying differences in batting context or underlying
hit probability.

### Alternative Null 2 — Context-Aware Bernoulli Null

Both the season-level Bernoulli and within-season permutation benchmarks
simplify the possibility that baseline hit probability changes across
at-bats. In actual game data, different at-bats may occur under different
baseball contexts.

A context-aware benchmark therefore allows the baseline hit probability
to vary across at-bats:

\[
H_i^* \sim \operatorname{Bernoulli}(\hat p_i),
\]

where

\[
\hat p_i = P(H_i = 1 \mid X_i)
\]

is estimated using contextual information available prior to the batting
outcome.

Simulated outcomes remain conditionally independent under this benchmark.
Therefore, systematic differences in baseline opportunity or difficulty
can be preserved while additional temporal dependence in batting outcomes
is removed.

Only variables that are available prior to the outcome are considered for
the baseline model in order to avoid outcome leakage.

```{r}
# names(ab_data)
```
*check the variables -- basline model*

```{r}
ab_data %>%
  summarise(
    n_stand = n_distinct(stand),
    n_p_throws = n_distinct(p_throws),
    n_home_team = n_distinct(home_team),
    n_away_team = n_distinct(away_team),
    n_pitcher = n_distinct(pitcher),
    n_inning = n_distinct(inning),
    n_outs = n_distinct(outs_when_up)
  )
```

*categorical distribution*

```{r}
ab_data %>%
  count(game_year, p_throws)

ab_data %>%
  count(game_year, stand)

ab_data %>%
  count(game_year, inning_topbot)
```

*missingness*

```{r}
ab_data %>%
  summarise(
    missing_p_throws = sum(is.na(p_throws)),
    missing_inning = sum(is.na(inning)),
    missing_outs = sum(is.na(outs_when_up)),
    missing_on_1b = sum(is.na(on_1b)),
    missing_on_2b = sum(is.na(on_2b)),
    missing_on_3b = sum(is.na(on_3b))
  )
```

#### construct pre-AB context variables


*creating the clean modeling dataset*

```{r}
context_data <- ab_data %>%
  arrange(
    game_year,
    game_date,
    game_pk,
    at_bat_number
  ) %>%
  mutate(
    pitcher_hand = factor(p_throws),

    home_away = factor(
      if_else(inning_topbot == "Bot", "Home", "Away")
    ),

    inning_group = case_when(
      inning <= 3 ~ "Early",
      inning <= 6 ~ "Middle",
      TRUE ~ "Late"
    ),
    inning_group = factor(inning_group),

    outs = factor(outs_when_up),

    runner_on_1b = as.integer(!is.na(on_1b)),
    runner_on_2b = as.integer(!is.na(on_2b)),
    runner_on_3b = as.integer(!is.na(on_3b))
  )
```

```{r}
context_data %>%
  summarise(
    n = n(),
    missing_hit = sum(is.na(hit)),
    missing_pitcher_hand = sum(is.na(pitcher_hand)),
    missing_home_away = sum(is.na(home_away)),
    missing_inning_group = sum(is.na(inning_group)),
    missing_outs = sum(is.na(outs)),
    missing_runner1 = sum(is.na(runner_on_1b)),
    missing_runner2 = sum(is.na(runner_on_2b)),
    missing_runner3 = sum(is.na(runner_on_3b))
  )
```

#### Context Model

A logistic regression model was used to estimate an at-bat-specific
baseline hit probability from pre-outcome contextual variables. The
model included season, pitcher handedness, home/away status, inning
group, outs, and base occupancy.

\[
\operatorname{logit}(p_i)
=
\alpha_{\text{season}(i)}
+
\beta^\top X_i.
\]

The purpose of this model is not to interpret individual coefficients
causally or to maximize predictive performance. Instead, it provides a
context-adjusted estimate of the baseline hit probability for each
official at-bat.

##### fit the context-aware logistic model

```{r}
context_model <- glm(
  hit ~ factor(game_year) +
    pitcher_hand +
    home_away +
    inning_group +
    outs +
    runner_on_1b +
    runner_on_2b +
    runner_on_3b,
  data = context_data,
  family = binomial(link = "logit")
)

summary(context_model)
```

##### generate p_hat_i

```{r}
context_data <- context_data %>%
  mutate(
    predicted_hit_prob = predict(
      context_model,
      newdata = context_data,
      type = "response"
    )
  )
```

##### check the predicted probabilities

```{r}
summary(context_data$predicted_hit_prob)
```
```{r}
context_data %>%
  summarise(
    min_p = min(predicted_hit_prob),
    q25_p = quantile(predicted_hit_prob, 0.25),
    median_p = median(predicted_hit_prob),
    mean_p = mean(predicted_hit_prob),
    q75_p = quantile(predicted_hit_prob, 0.75),
    max_p = max(predicted_hit_prob)
  )
```

##### sanity check?

compare the *observed season BA* VS *Average predicted probability*

```{r}
context_probability_check <- context_data %>%
  group_by(game_year) %>%
  summarise(
    n_ab = n(),
    observed_ba = mean(hit),
    mean_predicted_p = mean(predicted_hit_prob),
    min_predicted_p = min(predicted_hit_prob),
    max_predicted_p = max(predicted_hit_prob),
    .groups = "drop"
  )

context_probability_check
```

#### Context-Aware Monte Carlo Simulation

```{r}
simulate_context_variance <- function(
  predicted_prob,
  window = 25
) {

  simulated_hits <- rbinom(
    n = length(predicted_prob),
    size = 1,
    prob = predicted_prob
  )

  simulated_rolling_ba <- zoo::rollmean(
    simulated_hits,
    k = window,
    fill = NA,
    align = "right"
  )

  var(
    simulated_rolling_ba,
    na.rm = TRUE
  )
}
```

##### save every season's probability sequence

```{r}
context_sequences <- context_data %>%
  arrange(
    game_year,
    game_date,
    game_pk,
    at_bat_number
  ) %>%
  group_by(game_year) %>%
  summarise(
    predicted_prob = list(predicted_hit_prob),
    .groups = "drop"
  )
```

```{r}
context_sequences %>%
  rowwise() %>%
  mutate(
    n_probabilities = length(predicted_prob)
  ) %>%
  select(
    game_year,
    n_probabilities
  )
```

#### Context-Aware Monte Carlo Simulation

Using the fitted probabilities, 5,000 simulated seasons were generated
for each year according to

\[
H_i^* \sim \operatorname{Bernoulli}(\hat p_i).
\]

Thus, the simulated outcomes remain independent conditional on the
selected observed contexts, while the baseline probability of a hit is
allowed to vary across at-bats.

For each simulated season, the 25-at-bat rolling batting average and its
variance were recalculated. The observed variance was then compared with
the resulting context-aware null distribution using the same corrected
Monte Carlo p-value as in the primary analysis.

##### run 5000 context-aware simulations per season

```{r}
set.seed(123)

context_simulations <- tidyr::crossing(
  context_sequences,
  simulation = seq_len(n_sim)
) %>%
  rowwise() %>%
  mutate(
    simulated_variance =
      simulate_context_variance(
        predicted_prob = predicted_prob,
        window = 25
      )
  ) %>%
  ungroup()

context_simulations
```

```{r}
# check
context_simulations %>%
  count(game_year)
```

##### calculate context-aware monte carlo p-values

```{r}
context_results <- context_simulations %>%
  left_join(
    observed_primary_variance,
    by = "game_year"
  ) %>%
  group_by(game_year) %>%
  summarise(
    observed_variance =
      first(observed_variance),

    exceedances = sum(
      simulated_variance >= observed_variance
    ),

    n_simulations = n(),

    context_p = (
      exceedances + 1
    ) / (
      n_simulations + 1
    ),

    .groups = "drop"
  )

context_results
```

#### Context-Aware Monte Carlo Simulation

Using the fitted probabilities, 5,000 simulated seasons were generated
for each year according to

\[
H_i^* \sim \operatorname{Bernoulli}(\hat p_i).
\]

Thus, the simulated outcomes remain independent conditional on the
selected observed contexts, while the baseline probability of a hit is
allowed to vary across at-bats.

For each simulated season, the 25-at-bat rolling batting average and its
variance were recalculated. The observed variance was then compared with
the resulting context-aware null distribution using the same corrected
Monte Carlo p-value as in the primary analysis.


##  Comparison Across Null Models

```{r}
# Finalized primary results copied from main-analysis.Rmd.
# They are reused here to avoid rerunning the primary Monte Carlo test.

primary_bernoulli_results <- tibble::tibble(
  game_year = c(2022, 2023, 2024, 2025),

  observed_variance = c(
    0.006787898,
    0.009979931,
    0.010208230,
    0.014552027
  ),

  exceedances = c(
    3170,
    687,
    444,
    0
  ),

  n_simulations = rep(5000, 4),

  bernoulli_raw_p = c(
    0.63407319,
    0.13757249,
    0.08898220,
    0.00019996
  ),

  bernoulli_holm_p = c(
    0.63407319,
    0.27514497,
    0.26694661,
    0.00079984
  )
)

primary_bernoulli_results
```


```{r}
null_comparison <- primary_bernoulli_results %>%
  select(
    game_year,
    observed_variance,
    bernoulli_raw_p,
    bernoulli_holm_p
  ) %>%
  left_join(
    permutation_results %>%
      select(
        game_year,
        permutation_p
      ),
    by = "game_year"
  ) %>%
  left_join(
    context_results %>%
      select(
        game_year,
        context_p
      ),
    by = "game_year"
  )

null_comparison
```

Across the three null specifications, the results are highly consistent.
The 2022 and 2023 seasons do not exhibit unusually large 25-at-bat
rolling-BA variance under any of the three benchmarks. The 2024 season
shows moderately elevated variation, with p-values of approximately
0.09 across all three null models, but the observed statistic is not
especially extreme relative to the corresponding null distributions.

The 2025 season produces a substantially different pattern. Its observed
rolling-BA variance remains in the extreme upper tail under the
season-specific Bernoulli null (\(p \approx 0.0002\)), the within-season
permutation null (\(p \approx 0.0014\)), and the context-aware Bernoulli
null (\(p \approx 0.0008\)). The primary Bernoulli result also remained
statistically unusual after Holm adjustment across the four
season-specific tests (\(p_{\mathrm{Holm}} \approx 0.0008\)).

The close agreement across these alternative benchmarks indicates that
the 2025 result is not driven solely by allowing the simulated number of
hits to vary, by the constant season-level hit-probability assumption, or
by the selected observed pre-outcome contextual variables included in the
baseline model. However, these analyses establish robustness of the
temporal-variation finding rather than a causal hot-hand mechanism.
$$
\hat p_i = P(Y_i=1\mid X_i).
$$


# Robustness Summary



